Tidy data structure to support exploration and modeling of temporal-context data

### 2018-10-27 main.Rmd
described the mosaic plot and its data pipeline

The tsibble framework, as expected, incorporates this type of data, without the loss of data richness and heterogeneity. To declare the flight data as a valid tsibble, column `sched_dep_datetime` is specified as "index", columns `flight` and `origin` as "key" via `id(flight, origin)`. As a result of event data, this data is irregularly spaced, and hence switching `regular` to `FALSE` is necessary. The program internally validates if the key and index produce the distinct rows, and then sort the key and the index from past to recent. When the tsibble creation is done, the print display is data-oriented and contextually informative, such as dimensions, irregular interval (``r tsibble::tbl_sum(us_flights)[1]``) and the number of time-based observational units (``r tsibble::tbl_sum(us_flights)[2]``).

@@

Transforming tsibble for exploratory data analysis with a suite of time-specific and general-purpose manipulation verbs can generate sound pipelines. From the perspective of a passenger, one need to travel smart, by choosing an efficient carrier to fly with and the time of day to avoid congestion, for example. We take a drill-down approach to exploring this data, starting with annual carrier performance and disaggregating to finer time resolutions.

@@

(ref:carrier-mosaic-cap) Mosaic plot showing the association between the size of airline carriers and the delayed proportion of departure in 2017. Southwest Airlines is the largest operator, but does not operate as efficient as Delta. Hawaiian Airlines, also as a small operator, performs the best.

@@

```{r nyc-monthly-plot}
```

@@

(Need a flow chart for the code)

>>>>

The tsibble framework, as expected, incorporates this type of data, without the loss of data richness and heterogeneity. To declare the flight data as a valid tsibble, column `sched_dep_datetime` is specified as "index", columns `flight` and `origin` as "key" via `id(flight, origin)`. As a result of event data, this data is irregularly spaced, and hence switching to irregular option is necessary. The program internally validates if the key and index produce the distinct rows, and then sort the key and the index from past to recent. When the tsibble creation is done, the print display is data-oriented and contextually informative, such as dimensions, irregular interval (``r tsibble::tbl_sum(us_flights)[1]``) and the number of time-based observational units (``r tsibble::tbl_sum(us_flights)[2]``).

@@

Transforming tsibble for exploratory data analysis with a suite of time-specific and general-purpose manipulation verbs can generate well-constructed pipelines. From the perspective of a passenger, one need to travel smart, by choosing an efficient carrier to fly with and the time of day to avoid congestion, for example. We take a drill-down approach to exploring this data, starting with annual carrier performance and followed by disaggregating to finer time resolutions.

Figure \@ref(fig:carrier-mosaic) visually presents the end product of aggregating the number of on-time and delayed flights to the year interval by carriers. This pipeline is initialized defining a new variable if the flight is delayed, and involves summarizing the tallies of on-time and delayed flights for each carrier annually. To prepare the summarized data for a mosaic plot, it is further manipulated by melting new tallies into a single column. The flow chart shown as Figure \@ref(fig:flight-pipeline1) demonstrates the operations undertaking in a data pipeline. The input to this pipeline is a tsibble of irregular interval, and the output ends with a tsibble of unknown interval. The final data is each carrier along with a single year, in which the interval cannot be determined. It also feeds into the mosaic display. Southwest Airlines, as the largest carrier, operates less efficiently compared to Delta.

@@

(ref:carrier-mosaic-cap) Mosaic plot showing the association between the size of airline carriers and the delayed proportion of departure in 2017. Southwest Airlines is the largest operator, but does not operate as efficient as Delta. Hawaiian Airlines, also as a small operator, outperforms the rest.

```{r flight-pipeline1, fig.cap = "(ref:flight-pipeline1-cap)"}
include_graphics("img/flight-pipeline1")
```

(ref:flight-pipeline1-cap) Flow chart illustrates the pipeline that pre-processes the data for creating Figure \ref{fig:carrier-mosaic}.

@@

```{r nyc-monthly-plot, fig.cap = "(ref:nyc-monthly-plot-cap)"}
```

(ref:nyc-monthly-plot-cap) Daily departure percentage at New York airports.

@@
